---
title: "SoloProject"
format: html
---

The following map shows overall crime rates by county for all counties in Texas. The data is from TexasCounties.net who accessed the data from the Texas Department of Public Safety's Uniform Crime Reporting Section. The data is accessable from the following link: https://www.texascounties.net/statistics/crimerateoverall2020.htm

```{r Libraries}
library(readxl)
library(sf)
library(dplyr)
library(ggplot2)
library(stringr)
```

```{r Crime Data Load}
crime_data <- read_excel("/Users/owenlindstrom/Documents/2025/Stat 112/Solo Project.xlsx")
```

```{r Map Data Load}
tx_counties <- st_read("/Users/owenlindstrom/Documents/2025/Stat 112/Tx_CntyBndry_Jurisdictional_TIGER/Tx_CntyBndry_Jurisdictional_TIGER.shp")
```

```{r Clean Crime Data} 
crime_data_clean <- crime_data %>%
  mutate(subregion = tolower(County)) %>% 
  filter(CrimeRate < 7500)
```

```{r}
crime_data_clean <- crime_data_clean %>%
  mutate(County = str_to_title(County))

tx_counties_clean <- tx_counties %>%
  mutate(COUNTY = str_to_title(COUNTY))

```

```{r}
tx_counties_crime <- left_join(
  tx_counties_clean,
  crime_data_clean,
  by = c("COUNTY" = "County")
)

```

```{r}
ggplot(tx_counties_crime) +
  geom_sf(aes(fill = CrimeRate), color = "white", size = 0.2) +
  labs(
    title = "Texas Crime Rates by County",
    fill = "Crime Rate"
  ) +  scale_fill_gradient(low = "green2", high = "red")

```

For better visual context I dropped the two observations with the highest crime rates since they were large outliers in small counties that have a low population. When measuring Crime Rate per 10,000 people a small amount of crimes in these counties can create large estimates that create color schemes that don't effectively show the variation between other counties in crime rates. Crime rates are lowest in 2002 in the north central part of the state and highest on average in the south east part of the state and the central west part of the state. 
