# Import data
```{r}
hikes <- read.csv("https://mac-stat.github.io/data/high_peaks.csv")
```

##Excercise 1
```{r}
head(hikes)
```
a. The frequency of each catagory in the data set probably a bar chart
b. It would effectively show the spread of elevation across the mountains

##Excercise 2
```{r}
ggplot(hikes, aes(x = factor(rating, levels = c("easy", "moderate", "difficult")))) +
  geom_bar()
```


##Excercise 3
```{r}
count(hikes, rating)
```
1. 8 difficult, 11 easy, and 27 moderate hikes.
2. The hikes are not evenly distributed since their are more moderate hikes and less easy and difficult hikes.


##Excercise 4
```{r}
# 
ggplot(hikes, aes(x = rating)) +
  geom_bar()
```

```{r}
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar() +
  labs(x = "Rating", y = "Number of hikes")
```

```{r}
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar(fill = "blue") +
  labs(x = "Rating", y = "Number of hikes")
```

```{r}
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue") +
  labs(x = "Rating", y = "Number of hikes")
```

```{r}
# COMMENT on the change in the code and the corresponding change in the plot
ggplot(hikes, aes(x = rating)) +
  geom_bar(color = "orange", fill = "blue")  +
  labs(x = "Rating", y = "Number of hikes") +
  theme_minimal()
```


##Excercise 5


##Excercise 6


##Excercise 7


##Excercise 8


##Excercise 9


##Excercise 10


##Excercise 11


##Excercise 12


##Excercise 13
